---
title: "Predicting Housing Prices Using Multiple Regression"
author:
  - Group: "G06D02"
  - Name: "Jonah Wee"
  - Name: "Kok Yuan"
  - Name: "Zhe Kai"
  - Name: "Jian Xin"
format:
  html:
    theme: default
    toc: false
    embed-resources: true
  pdf:
    documentclass: article
    geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  fig.align = "center"
)

library(tidyverse)
library(GGally)
library(broom)
```

## Abstract

This study investigates which property characteristics most influence residential
sale prices using data from the Saratoga, New York housing market. After cleaning
1,736 observations and applying stepwise variable selection, a multiple regression
model with ten predictors was fit to 1,721 transactions. The final model explains
approximately 65% of the variation in sale price (adjusted R² = 0.645). Living
area, land value, and bathrooms emerged as the strongest positive drivers of price,
while waterfront access carried the single largest premium at roughly \$120,000.
The model provides useful insight into housing value drivers but is limited by
right-skewed residuals and the absence of neighbourhood-level covariates.

## Introduction

What property characteristics most influence residential sale prices? Answering
this question has practical value for home buyers evaluating fair market value,
sellers setting listing prices, and appraisers performing valuations. This report
builds a multiple linear regression model to predict sale price from structural,
locational, and categorical property features recorded in the Saratoga housing
dataset.

## Data Set

```{r load-data}
train <- read.csv("cleaned_housing_data.csv")

# Convert binary and categorical variables to factors
train <- train %>%
  mutate(
    Waterfront    = factor(Waterfront),
    New.Construct = factor(New.Construct),
    Central.Air   = factor(Central.Air),
    Fuel.Type     = factor(Fuel.Type),
    Heat.Type     = factor(Heat.Type),
    Sewer.Type    = factor(Sewer.Type)
  )
```

The dataset records residential property sales from Saratoga, New York and is
commonly used for regression instruction. The raw file contained 1,736
observations across 17 variables. The administrative `Test` column — a binary
partition indicator — was excluded prior to analysis, leaving 16 predictors and
one response variable, **Price** (USD).

**Data cleaning.** One duplicate row was removed, as exact matches across all
16 variables more plausibly reflect data duplication than distinct transactions.
Six additional observations were excluded for implausible values: three where
`Price < Land.Value` (implying negative structural value) and three with
zero-valued structural variables (`Lot.Size = 0` or `Bathrooms = 0`). The
final cleaned dataset contains `r nrow(train)` observations and no missing
values.

**Variables.** Predictors span continuous measures (Lot.Size, Age, Land.Value,
Living.Area, Pct.College, Bedrooms, Fireplaces, Bathrooms, Rooms), binary
indicators (Waterfront, New.Construct, Central.Air), and categorical factors
(Fuel.Type, Heat.Type, Sewer.Type). Four sparse Fuel.Type levels (each n ≤ 3)
were collapsed into "Other."

```{r summary-table}
train %>%
  select(Price, Living.Area, Land.Value, Age, Lot.Size, Bathrooms, Rooms) %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(Mean = mean(value), SD = sd(value),
            Min = min(value), Max = max(value), .groups = "drop") %>%
  knitr::kable(digits = 0, caption = "Summary statistics for key variables.")
```

```{r eda-plots, fig.height=3, fig.cap="Scatter plots of two key continuous predictors against Price."}
p1 <- ggplot(train, aes(x = Living.Area, y = Price)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_y_continuous(labels = scales::dollar) +
  labs(x = "Living Area (sq ft)", y = "Price (USD)") +
  theme_minimal(base_size = 9)

p2 <- ggplot(train, aes(x = Land.Value, y = Price)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_continuous(labels = scales::dollar) +
  labs(x = "Land Value (USD)", y = "Price (USD)") +
  theme_minimal(base_size = 9)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Analysis

**Variable selection.** A full model with all 15 predictors was fit first.
Stepwise selection using AIC (direction = "both") was then applied. Both
forward and backward procedures converged on the same ten-predictor model, so
only the final model is reported. The dropped variables — Fuel.Type, Heat.Type,
Sewer.Type, Pct.College, and Fireplaces — were individually and jointly
non-significant (all p > 0.08 in the full model).

```{r variable-selection}
full_model <- lm(Price ~ ., data = train)
step_model <- step(full_model, direction = "both", trace = 0)
```

**Assumptions.** Each observation represents a distinct property transaction,
so the independence assumption (that rows are independent of one another) is
reasonable. Residual diagnostic plots are shown below.

```{r residual-diagnostics, fig.height=5.5, fig.width=6, fig.cap="Residual diagnostic plots for the final model."}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(step_model)
```

The residuals-versus-fitted plot shows no strong curvature, supporting the
linearity assumption. However, the spread of residuals fans outward at higher
fitted values, indicating some heteroscedasticity, and the Q–Q plot reveals
right-skew in the upper tail. These departures suggest that coefficient
estimates remain unbiased but standard errors may be somewhat understated for
the highest-priced homes. A log transformation of Price was explored but
altered the set of significant predictors and complicated coefficient
interpretation; the untransformed model was retained for clarity, and the
residual non-normality is noted as a limitation.

## Results

```{r results-table}
knitr::kable(
  tidy(step_model, conf.int = TRUE) %>%
    select(term, estimate, std.error, p.value, conf.low, conf.high),
  digits = 2,
  col.names = c("Term", "Estimate", "SE", "p-value", "CI Low", "CI High"),
  caption = "Final model coefficient estimates with 95% confidence intervals."
)
```

```{r model-fit}
glance_stats <- glance(step_model)
```

The final model has an adjusted R² of `r round(glance_stats$adj.r.squared, 3)`,
a residual standard error of \$`r format(round(glance_stats$sigma, 0), big.mark = ",")`,
and all ten retained predictors are significant at α = 0.05.

**Key coefficient interpretations** (holding other predictors constant):

- **Living.Area:** each additional 100 sq ft of living space is associated with
  an approximately \$`r format(round(coef(step_model)["Living.Area"] * 100, 0), big.mark = ",")` increase in price (range in data: `r min(train$Living.Area)` – `r format(max(train$Living.Area), big.mark = ",")` sq ft).
- **Land.Value:** a \$10,000 increase in assessed land value corresponds to
  roughly \$`r format(round(coef(step_model)["Land.Value"] * 10000, 0), big.mark = ",")` higher sale price.
- **Bathrooms:** each additional bathroom adds about \$`r format(round(coef(step_model)["Bathrooms"], 0), big.mark = ",")` (range: 1–4.5).
- **Waterfront:** waterfront properties sell for approximately
  \$`r format(round(coef(step_model)["Waterfront1"], 0), big.mark = ",")` more,
  though this estimate is based on only `r sum(train$Waterfront == "1")` waterfront
  homes and should be interpreted with caution.
- **New.Construct:** newly constructed homes are associated with a
  \$`r format(abs(round(coef(step_model)["New.Construct1"], 0)), big.mark = ",")`
  *lower* price after controlling for other features, likely because new builds in
  this market tend to occupy lower-valued land.
- **Bedrooms:** an additional bedroom is associated with a
  \$`r format(abs(round(coef(step_model)["Bedrooms"], 0)), big.mark = ",")`
  *decrease* in price once living area and rooms are held constant, reflecting the
  trade-off between more bedrooms and smaller room sizes within a fixed footprint.

## Discussion & Conclusion

This model explains roughly 65% of the variance in Saratoga housing prices.
Living area, land value, and bathrooms are the strongest positive value drivers,
while waterfront access carries the single largest price premium. These findings
are consistent with conventional real-estate intuition: space, location quality,
and desirable amenities command higher prices.

**Limitations.** The data are observational, so no definite claims can be made. The
right-skewed residuals and mild heteroscedasticity mean that predictions for
the most expensive homes carry wider uncertainty than the model implies.
Additionally, potentially important predictors such as school district,
neighbourhood desirability, and property condition are not captured. The dataset
is specific to Saratoga, NY, and the model may not transfer to other housing
markets with different price distributions or regional preferences.

**Overcoming limitations.** Future work could incorporate spatial variables
(e.g. distance to amenities, school ratings) and apply robust standard errors
or a log transformation to better handle residual non-normality. Expanding the
dataset across multiple markets would improve generalisability.

**Conclusion.** A ten-predictor linear regression provides a useful, interpretable
summary of the factors driving housing prices in Saratoga. While not a substitute
for a professional appraisal, the model identifies living area, land value,
bathrooms, and waterfront access as the most influential value determinants and
can serve as a starting point for more refined pricing tools.

{{< pagebreak >}}

## Appendix

```{r appendix-stepwise-summary}
cat("Stepwise model formula:\n")
print(formula(step_model))
cat("\n")
summary(step_model)
```

```{r appendix-full-model-aic}
cat("Full model AIC:", AIC(full_model), "\n")
cat("Step model AIC:", AIC(step_model), "\n")
```
